---
title: "STA2201 Homework One(Lab One)"
author: shaohan chang
output:
  html_document: default
  pdf_document: default
date: "2023-01-13"
---
# Lab Exercises


Explanation: Download the packages of tidyverse.
```{r}
#| message: false
#install.packages("tidyverse")
library(tidyverse)

```


Explanation: download the table from web-link.
```{r}
dm <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", skip = 2, col_types = "dcddd")
head(dm)

```


  
#1. Plot the ratio of male to female mortality rates over time for ages 10,20,30 and 40 (different color for each age) and change the theme 

```{r}
ratio <- dm |> 
  filter(Age== 10 |Age==20 | Age == 30 | Age == 40) |> 
  mutate(Male_to_Female_Mortality = Male/Female)


ratio |> 
  ggplot(aes(x = Year , y = Male_to_Female_Mortality ,color = Age)) +
  geom_line() + 
  scale_color_brewer(palette = 'Set1')
```


#2. Find the age that has the highest female mortality rate each year 

Explanation: Here is the list of the age that has the highest felmale mortality rate each year, however, the data result is not suprise, all of the result in highest felmale mortality rate each year, age are higher than 100, which is in my result of expectation. 

```{r}

group_by_year <- dm |>
  group_by(Year)|>
  summarise(highest_female_mortality_rate=Age[which.max(Female)])


group_by_year


```


#3. Use the `summarize(across())` syntax to calculate the standard deviation of mortality rates by age for the Male, Female and Total populations.

Explanation:Used the method of summarize(across()) to calculate the standard deviation of mortality rates by age for the Male, Female and Total populations.

```{r}

rate_of_Male <- dm |>
  group_by(Age)|>
  summarize(across(Total:Male|Female,sd,na.rm = TRUE)) 
rate_of_Male
  

```



#4. The Canadian HMD also provides population sizes over time (https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt). Use these to calculate the population weighted average mortality rate separately for males and females, for every year. Make a nice line plot showing the result (with meaningful labels/titles) and briefly comment on what you see (1 sentence). Hint: `left_join` will probably be useful here. 

Explanation:read the table from web-link

```{r}
pp <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt", skip = 2, col_types = "dcddd")
head(pp)

```

Explanation: AS showed in the figure,In overall,the mortality rate of population weighted average for females and males from 1921 to 2020 showed an increasing trending. 


```{r}

weighted_mean_table_female <- pp |>
  group_by (Year) |>
  summarize(weighted_average_female = weighted.mean(Female,Total,na.rm = TRUE))


weighted_mean_table_male <- pp |>
  group_by (Year) |>
  summarize(weighted_average_male = weighted.mean(Male,Total,na.rm = TRUE))

fix_up = left_join(weighted_mean_table_female,weighted_mean_table_male, by='Year')

fix_up <- fix_up |>
  select(Year,weighted_average_female,weighted_average_male) |>
  pivot_longer(weighted_average_female:weighted_average_male, names_to = "Sex", values_to = "Mortality")

fix_up |>
  ggplot(aes(x = Year, y = Mortality, color = Sex, linetype = Sex)) + 
  geom_line() + 
  scale_color_brewer(palette = "Set1") +
  ggtitle("Mortality Rate of Population Weighted Average for Females and Males from 1921 to 2020") +
  theme(plot.title = element_text(size = 8, color = "black", face = "bold"))
```