---
title: "Week 6: Visualizing the Bayesian Workflow"
author: 'SHAOHAN CHANG'
date: today
date-format: "DD/MM/YY"
format: pdf
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Introduction

This lab will be looking at trying to replicate some of the
visualizations in the lecture notes, involving prior and posterior
predictive checks, and LOO model comparisons.

The dataset is a 0.1% of all births in the US in 2017. I've pulled out a
few different variables, but as in the lecture, we'll just focus on
birth weight and gestational age.

# The data

Read it in, along with all our packages.

```{r}
library(tidyverse)
library(here)
# for bayes stuff
library(rstan)
library(bayesplot) 
library(loo) 
library(tidybayes) 
#ds <- readRDS('births_2017_sample.RDS')
ds <- readRDS('C:/Users/admin/Desktop/Lab6/births_2017_sample.RDS')
head(ds)
```

Brief overview of variables:

-   `mager` mum's age
-   `mracehisp` mum's race/ethnicity see here for codes:
    https://data.nber.org/natality/2017/natl2017.pdf page 15
-   `meduc` mum's education see here for codes:
    https://data.nber.org/natality/2017/natl2017.pdf page 16
-   `bmi` mum's bmi
-   `sex` baby's sex
-   `combgest` gestational age in weeks
-   `dbwt` birth weight in kg
-   `ilive` alive at time of report y/n/ unsure

I'm going to rename some variables, remove any observations with missing
gestational age or birth weight, restrict just to babies that were
alive, and make a preterm variable.

```{r}
ds <- ds |> 
  rename(birthweight = dbwt, gest = combgest) |> 
  mutate(preterm = ifelse(gest<32, "Y", "N")) |> 
  filter(ilive=="Y",gest< 99, birthweight<9.999)
```

## Question 1

Use plots or tables to show three interesting observations about the
data. Remember:

-   Explain what your graph/ tables show
-   Choose a graph type that's appropriate to the data type
-   If you use `geom_smooth`, please also plot the underlying data

Feel free to replicate one of the scatter plots in the lectures as one
of the interesting observations, as those form the basis of our models.

Answer: I choose the first plot,since it is the well represent the
relationship of the birth weight and the gestation age , and the
different color of the preterm status. Which is clear to show the
different relationship.

One:

I make a scatter plot using ggplot to find the relationship between the
Birth weight and the Gestational age. As the description given above,
the statute of the preterm is remark as Y and N. Which can clearly to
show the relationship between the birth weight and the gestational age
in different status of preterm status.

```{r warning=FALSE}
plot_one <- ds |> 
  group_by(preterm)


ggplot(data = plot_one, aes(x = gest, y = birthweight, color = preterm)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Scatterplot of Birthweight by Gestational Age",
       x = "Gestational Age (weeks)", y = "Birthweight (grams)",
       color = "Preterm Status") +
  scale_color_viridis_d() +
  theme_classic()
```

Two: I use ggplot to see if mother's BMI is related to birth weight.
From the plot, I find there is no people with BMI of 100, which suggest
that the data may not very reliable.A data visualization package in R,
to investigate the relationship between a mother's BMI and the birth
weight of their child. Created a scatter plot that displays the mother's
BMI on the x-axis and the birth weight on the y-axis.

```{r}
ds |>
  ggplot() +
  aes(x = bmi, y = birthweight, color = preterm) +
  geom_smooth(method = "lm", se = FALSE, size = 1, color = "black") +
  geom_point(alpha = 0.5, shape = 19) +
  scale_color_manual(values = c("blue", "green"), name = "Preterm") +
  theme_classic() +
  labs(title = "Mother's BMI vs Baby Birthweight",
       x = "Mother's BMI (kg/m²)",
       y = "Baby Birthweight (g)") +
  theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
        axis.title = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 14))

```

Three:

```{r}
levels_race=c("NHW","NHB","NHAIAN","NHA","NHOPI","Hisp >1 race","Hisp","Unknown")
library(forcats)
ds$race=as_factor(ds$mracehisp)
levels(ds$race)=levels_race

tab1=ds|>group_by(race,sex) |>summarize(n = n(),mean_gest = mean(gest),
med_gest = median(gest),var_gest = var(gest),
mean_bwirthweight = mean(birthweight),var_bwirthweight = median(birthweight),
sum_preterm = sum(preterm == "Y"),
prop_preterm = mean(preterm == "Y")) |> arrange(n)

tab1

```

The tibble previously mentioned gives an overview of gestational age and
birth weight for various racial and sexual combinations. In order to
account for variations in gestational age and birth weight, this
information can be used to decide whether race and sex should be
included as factors.According to the tibble, there does not appear to be
a significant difference between the various groups in the percentage of
preterm births or birth weight. In contrast to the other groups,
Non-Hispanic Black (NHB) infants have a higher variance of gestational
age, with the exception of situations where the sample size is
small.This information suggests that additional research may be required
to determine whether race and gender may have an impact on gestational
age and birth weight. It's conceivable that adding these variables to a
statistical analysis as covariates could aid in providing a more
comprehensive explanation for any observed differences or variability in
these results between groups. The specific research question being
examined and the analysis's objectives, however, would ultimately
determine this choice.

# The model

As in lecture, we will look at two candidate models

Model 1 has log birth weight as a function of log gestational age

$$
\log(y_i) \sim N(\beta_1 + \beta_2\log(x_i), \sigma^2)
$$

Model 2 has an interaction term between gestation and prematurity

$$
\log(y_i) \sim N(\beta_1 + \beta_2\log(x_i) + \beta_3 z_i + \beta_4\log(x_i) z_i, \sigma^2)
$$

-   $y_i$ is weight in kg
-   $x_i$ is gestational age in weeks, CENTERED AND STANDARDIZED
-   $z_i$ is preterm (0 or 1, if gestational age is less than 32 weeks)

# Prior predictive checks

Let's put some weakly informative priors on all parameters i.e. for the
$\beta$s

$$
\beta \sim N(0, 1)
$$

and for $\sigma$

$$
\sigma \sim N^+(0,1)
$$ where the plus means positive values only i.e. Half Normal.

Let's check to see what the resulting distribution of birth weights look
like given Model 1 and the priors specified above, assuming we had no
data on birth weight (but observations of gestational age).

## Question 2

For Model 1, simulate values of $\beta$s and $\sigma$ based on the
priors above. Do 1000 simulations. Use these values to simulate (log)
birth weights from the likelihood specified in Model 1, based on the set
of observed gestational weights. **Remember the gestational weights
should be centered and standardized**.

-   Plot the resulting distribution of simulated (log) birth weights.
-   Plot ten simulations of (log) birthweights against gestational age.

```{r}
# we know which is the 3842 row in the data of ds. 
nrow(ds)
```

```{r}

ites = 1000
beta0 = rnorm(ites)
beta1 = rnorm(ites)
sigma = abs(rnorm(ites))
log_gest = scale(log(ds$gest))

log_birthweight = matrix(0,ncol=ites,nrow=3842)

for (i in 1:ites) {
  set.seed(i+3842)
  mu <- beta0[i] + beta1[i]*log_gest
  log_birthweight[,i]=rnorm(3842, mean = mu, sd = sigma[i])
}
log_birthweight=as.data.frame(log_birthweight)

# I set the bins is 90 (scale)
library(ggplot2)
library(tidyr)

log_birthweight|>pivot_longer(cols = everything())|> ggplot(aes(x = value))+geom_histogram(bins = 90)+
  theme_bw() + scale_x_continuous(name = "log_birth_weight")


# The code that I did the ten simulations of the (log) birthweight against gestational age. 

sim10 <- log_birthweight[,1:10]

for(i in 1:10){
  logbirth<-sim10[,i]
  dmatric<- as.data.frame(cbind(log_gest,logbirth))
 print( ggplot(data = dmatric, aes(x = log_gest, y = logbirth)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "(log) birthweights against gestational age",
       x = "gestational age", y = "(log) birthweights"))
}

```

# Run the model

Now we're going to run Model 1 in Stan. The stan code is in the
`code/models` folder.

First, get our data into right form for input into stan.

```{r}
ds$log_weight <- log(ds$birthweight)
ds$log_gest_c <- (log(ds$gest) - mean(log(ds$gest)))/sd(log(ds$gest))
# put into a list
stan_data <- list(N = nrow(ds),
                  log_weight = ds$log_weight,
                  log_gest = ds$log_gest_c)
```

Now fit the model

```{r}
mod1 <- stan(data = stan_data, 
             file = here("C:/Users/admin/Desktop/Lab6/simple_weight.stan"),
             iter = 500,
             seed = 243)

```

```{r}
summary(mod1)$summary[c("beta[1]", "beta[2]", "sigma"),]
```

## Question 3

Based on model 1 , give an estimate of the expected birthweight of a
baby who was born at a gestational age of 37 weeks.

```{r}
library(tidybayes)
mod1|>
  spread_draws(beta[condition], sigma)|>
  mutate(condition_mean = exp( beta[1] + beta[2]*log(37))) |>
  median_qi(condition_mean)

```

## Question 4

Write a stan model to run Model 2, and run it.

```{r}
preterm = ifelse(ds$preterm == "Y", 1, 0)

stan_data[["preterm"]] = preterm

stan_data[["interac"]] = preterm*log(ds$gest)

mod2 <- stan(data = stan_data,
  file = here("C:/Users/admin/Desktop/Lab6/simple_weight_interaction.stan"),
  iter = 500,
  seed = 243)
```

```{r}
# I need to summary the beta from 1 to 4. 
summary(mod2)$summary[c(paste0("beta[", 1:4, "]"), "sigma"),]
```

## Question 5

For reference I have uploaded some model 2 results. Check your results
are similar.

```{r}
# load the mod2 file.
load(here("C:/Users/admin/Desktop/Lab6/mod2.Rda"))
summary(mod2)$summary[c(paste0("beta[", 1:4, "]"), "sigma"),]

```

# PPCs

Now we've run two candidate models let's do some posterior predictive
checks. The `bayesplot` package has a lot of inbuilt graphing functions
to do this. For example, let's plot the distribution of our data (y)
against 100 different datasets drawn from the posterior predictive
distribution:

```{r}
set.seed(1856)
y <- ds$log_weight
yrep1 <- extract(mod1)[["log_weight_rep"]]
yrep2 <- extract(mod2)[["log_weight_rep"]] 
dim(yrep1)
samp100 <- sample(nrow(yrep1), 100)
ppc_dens_overlay(y, yrep1[samp100, ])  + ggtitle("distribution of observed versus predicted birthweights")
```

## Question 6

Make a similar plot to the one above but for model 2, and **not** using
the bayes plot in built function (i.e. do it yourself just with
`geom_density`)

```{r}
set.seed(200)
draw_samples2 = sample(nrow(yrep2), 100)
tab2 = as_tibble(t(yrep2[draw_samples2, ]))

colnames(tab2) <- 1:100
df2 = as_tibble(tab2) 
density2 = df2 |> # amount of 3842
  bind_cols(n = 1:3842, log_weight_obs = log(ds$birthweight)) |>
  pivot_longer(`1`:`100`, names_to = "simulation", values_to = "log_weight_rep")

# plot densities 

density2 |> ggplot(aes(log_weight_rep, group = simulation)) +
geom_density( aes(color = "y_rep")) +
geom_density(data = ds |> mutate(simulation= 1),
aes(x =log(birthweight), col = "y")) +
scale_color_manual(name = "",values = c("y" = "blue","y_rep" = "red")) +
ggtitle("Birthweight distributions as compared to predictions") +
theme_bw(base_size = 13) 

```

## Test statistics

We can also look at some summary statistics in the PPD versus the data,
again either using `bayesplot` -- the function of interest is `ppc_stat`
or `ppc_stat_grouped` -- or just doing it ourselves using ggplot.

E.g. medians by prematurity for Model 1

```{r}
ppc_stat_grouped(y, yrep1, group = ds$preterm, stat = 'median')
```

## Question 7

Use a test statistic of the proportion of births under 2.5kg. Calculate
the test statistic for the data, and the posterior predictive samples
for both models, and plot the comparison (one plot per model).

```{r}
# I set the true value.
true_value =  mean(ds$birthweight < 2.5)
library(tidyverse)
#set the test rep.
test_rep = density2 |> 
     group_by(simulation) |>
  summarize(test_stat = mean(exp(log_weight_rep) < 2.5))

test_rep |> ggplot(aes(x = test_stat)) +
geom_histogram(bins = 20, fill = "red") +
geom_vline(xintercept = true_value, color = "blue") +
theme_bw() +
labs(caption = "real test statistic in blue and simulated test statistic in red",
title = "Test Results for PPC for the percentage of bw under 2.5 kg")

```

# LOO

Finally let's calculate the LOO elpd for each model and compare. The
first step of this is to get the point-wise log likelihood estimates
from each model:

```{r}
loglik1 <- extract(mod1)[["log_lik"]]
loglik2 <- extract(mod2)[["log_lik"]]
```

And then we can use these in the `loo` function to get estimates for the
elpd. Note the `save_psis = TRUE` argument saves the calculation for
each simulated draw, which is needed for the LOO-PIT calculation below.

```{r}
loo1 <- loo(loglik1, save_psis = TRUE)
loo2 <- loo(loglik2, save_psis = TRUE)
```

Look at the output:

```{r}
loo1
loo2
```

Comparing the two models tells us Model 2 is better:

```{r}
loo_compare(loo1, loo2)
```

We can also compare the LOO-PIT of each of the models to standard
uniforms. The both do pretty well.

```{r}
ppc_loo_pit_overlay(yrep = yrep1, y = y, lw = weights(loo1$psis_object))
ppc_loo_pit_overlay(yrep = yrep2, y = y, lw = weights(loo2$psis_object))
```

## Bonus question (not required)

Create your own PIT histogram "from scratch" for Model 2.

```{r}
library(scoringutils)
pit <- pit_sample(y,t(yrep2))
plot_pit(pit)


```

## Question 8

Based on the original dataset, choose one (or more) additional
covariates to add to the linear regression model. Run the model in Stan,
and compare with Model 2 above on at least 2 posterior predictive
checks.

Model 3 adds "mager" to the linear regression model (model 2).

$$
\log(y_i) \sim N(\beta_1 + \beta_2\log(x_i)+ \beta_3\log(z_i)+\beta_4\log(x_i)z_i+\beta_5\log(m_i), \sigma^2)
$$ - $y_i$ is weight in kg - $x_i$ is gestational age in weeks, CENTERED
AND STANDARDIZED - $z_i$ is preterm - $m_i$ is mum' age

```{r}
# put into a list
ds$log_age=log(ds$mager)
preterm <- ifelse(ds$preterm == "Y", 1, 0)
log_weight = log(ds$birthweight)
log_gest = log(ds$gest)
log_age<-ds$log_age
interac<- preterm*log(ds$gest)
stan_data1 <- list(N = nrow(ds),
                  log_weight=log_weight,
                  log_gest=log_gest,
                  preterm=preterm,
                  log_age=log_age,interac=interac)


mod3<- stan(data = stan_data1,
  file = here("C:/Users/admin/Desktop/Lab6/simple_weight_mod3.stan"),
  iter = 500,
  seed = 200)


sum_mod3 = summary(mod3)
sum_mod3$summary[c(paste0("beta[", 1:5, "]"), "sigma"),]



set.seed(300)
y <- ds$log_weight
yrep2<- extract(mod2)[["log_weight_rep"]]
yrep3<- extract(mod3)[["log_weight_rep"]]
draw_samples3<- sample(nrow(yrep3), 100)
ppc_dens_overlay(y, yrep3[draw_samples3, ]) + ggtitle("Distribution of observed versus predicted birthweight from Model 3")

draw_samples2 <- sample(nrow(yrep2), 100)
ppc_dens_overlay(y, yrep2[draw_samples2, ]) + ggtitle("Distribution of observed versus predicted birthweight from Model 2")



ppc_stat_grouped(ds$log_weight, yrep2, group = ds$preterm, stat = 'median')
ppc_stat_grouped(ds$log_weight, yrep3, group = ds$preterm, stat = 'median')

loglik3 <- extract(mod3)[["log_lik"]]
loglik2 <- extract(mod2)[["log_lik"]]
loo3<- loo(loglik3, save_psis = TRUE)
loo2 <- loo(loglik2, save_psis = TRUE)


compare(loo3, loo2)
ppc_loo_pit_overlay(yrep = yrep2, y = y, lw = weights(loo2$psis_object))
ppc_loo_pit_overlay(yrep = yrep3, y = y, lw = weights(loo3$psis_object))
ppc_loo_pit_overlay(yrep = yrep3, y = y, lw = weights(loo3$psis_object))

```
